---
title: "Loan Default & Loss Prediction — ADM Group Project"
author: "Group 5 — Shreeya Pathak, Sri Anu Vunnam, Inn Kyung Seo"
date: "2026-05-01"
output: word_document
---
## Group Members & Contributions

| Name | Contribution |
|---|---|
| Shreeya Pathak | Data preprocessing, Feature selection, Report writing |
| Sri Anu Vunnam | Model building, XGBoost tuning, Predictions |
| Inn Kyung Seo | EDA, Random Forest modeling, Presentation |

## Project Overview

The banks and financial institutions will lose **millions** of dollars every year because of the issue of loan defaulting. To be able to know which of the loans will default and how much financial loss there will be from it, becomes very crucial for the lenders. This project will use machine learning to accomplish that exact thing.

### Objectives
- Prediction of whether the loan will **default or not**
- Prediction of the amount of **financial loss**
- Model evaluation based on **Mean Absolute Error (MAE)**

### Data Set
- Training dataset consists of 80,000 loans
- Total of 763 features per each loan
- Target value refers to loss amount

### Two-Step Approach
- Step 1: Classification of loan defaulting (Yes or No)
- Step 2: Prediction of loss amount

## Setup & Libraries
During this stage, we will import all the required libraries along with setting the file path to our training and test datasets. We have also assigned a random seed value because we need to make sure that the output is replicable each time we execute the program.

```{r setup, message=FALSE, warning=FALSE}
# File Paths
TRAIN_PATH <- "C:/Users/sanju/Downloads/train_v3.csv"
TEST_PATH  <- "C:/Users/sanju/Downloads/test__no_lossv3.csv"

# Load required libraries
required_pkgs <- c("readr", "dplyr", "ggplot2", "caret", "glmnet",
                   "randomForest", "xgboost", "pROC", "corrplot")

new_pkgs <- required_pkgs[!(required_pkgs %in% installed.packages()[,"Package"])]
if (length(new_pkgs) > 0) install.packages(new_pkgs, dependencies = TRUE)

library(readr)
library(dplyr)
library(ggplot2)
library(caret)
library(glmnet)
library(randomForest)
library(xgboost)
library(pROC)
library(corrplot)

set.seed(123)
```

## Loading Data
At this stage, we import the dataset for training in R. The training dataset consists of actual loan information and the loss amount, which our model learns from. At the same time, we have an initial glance at the data set.

```{r load-data, message=FALSE}
# Load training data
raw_data <- read_csv(TRAIN_PATH, show_col_types = FALSE)

# Basic overview
cat("TRAINING DATA OVERVIEW\n")
cat(sprintf("Rows         : %d\n", nrow(raw_data)))
cat(sprintf("Columns      : %d\n", ncol(raw_data)))
cat(sprintf("Defaulted    : %d  (%.2f%%)\n",
            sum(raw_data$loss > 0, na.rm = TRUE),
            mean(raw_data$loss > 0, na.rm = TRUE) * 100))
cat(sprintf("Non-defaulted: %d  (%.2f%%)\n",
            sum(raw_data$loss == 0, na.rm = TRUE),
            mean(raw_data$loss == 0, na.rm = TRUE) * 100))
```
As per the result displayed above, the number of loans in the training set is **80,000** and features amount to **763**. Out of those 80,000 loans, **7,379 loans** had defaulted, which means a default rate of **9.22%**, while non-default loans were **72,621 (90.78%)**.

## Exploratory Data Analysis (EDA)
This section discusses visual explorations of the data for deeper insights into its nature, relationships and quality which will help us make appropriate decisions in terms of data preprocessing and modeling.

### Plot 1 — Default vs Non-Default Distribution

This chart depicts the percentage of **default and non-default loans** in the dataset. It gives us a sense of the problem of imbalanced classes and helps us choose the proper modeling technique.
```{r eda-plot1, message=FALSE, warning=FALSE}
plot_data <- raw_data %>%
  mutate(loan_status = ifelse(loss > 0, "Defaulted", "Non-Defaulted")) %>%
  count(loan_status) %>%
  mutate(pct = round(n / sum(n) * 100, 1),
         label = paste0(loan_status, "\n", format(n, big.mark=","), "\n(", pct, "%)"))

ggplot(plot_data, aes(x = "", y = n, fill = loan_status)) +
  geom_col(width = 1, color = "white") +
  coord_polar("y") +
  geom_text(aes(label = label), position = position_stack(vjust = 0.5), size = 4) +
  scale_fill_manual(values = c("Defaulted" = "#E05C5C", "Non-Defaulted" = "#5C9BE0")) +
  labs(title = "Loan Default Distribution") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        legend.position = "none")
```
The graph clearly shows that the data set is extremely imbalanced. The data set consists of 80,000 loans, out of which **7,379 loans defaulted**. This means that only **9.22%** of the total population defaulted, while the rest **72,621 loans or 90.78%** did not default. The imbalance in the dataset suggests that we should pay attention to our model training.

### Plot 2 — Loss Distribution for Defaulted Loans

This graph depicts the distribution of the amounts of losses from the defaulted loans. This can give us insight into the magnitude of the losses and whether the loss values follow a **normal or skewed distribution.**
```{r eda-plot2, message=FALSE, warning=FALSE}
defaulted_only <- raw_data %>% filter(loss > 0)

ggplot(defaulted_only, aes(x = loss)) +
  geom_histogram(bins = 50, fill = "#E05C5C", color = "white", alpha = 0.85) +
  geom_vline(xintercept = mean(defaulted_only$loss), color = "blue", linetype = "dashed", linewidth = 1) +
  geom_vline(xintercept = median(defaulted_only$loss), color = "green", linetype = "dashed", linewidth = 1) +
  labs(title = "Loss Distribution for Defaulted Loans",
       x = "Loss Amount", y = "Count") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```
The graph clearly shows that losses from defaulted loans are extremely **right skewed**. The majority of the loss amounts on defaulted loans concentrate within the **0 to 25 **range. Nevertheless, some loan loss amounts as high as 100 are the outliers. **Median is the green dashed line while mean is the blue dashed line**. The fact that mean is greater than median confirms the right skewness of the data.

### Plot 3 — Missing Value Analysis
This plot shows how many missing values exist in our dataset and how severe they are across different columns. Identifying missing values is important because most machine learning models cannot handle them directly and we need to decide how to deal with them before modeling.

```{r eda-plot3, message=FALSE, warning=FALSE}
col_missing_pct <- colMeans(is.na(raw_data)) * 100

missing_summary <- data.frame(
  Category = c("No Missing (0%)", "Low (0-10%)", "Medium (10-50%)", "High (>50%)"),
  Count = c(
    sum(col_missing_pct == 0),
    sum(col_missing_pct > 0 & col_missing_pct <= 10),
    sum(col_missing_pct > 10 & col_missing_pct <= 50),
    sum(col_missing_pct > 50)
  )
)

ggplot(missing_summary, aes(x = Category, y = Count, fill = Category)) +
  geom_col(color = "white", show.legend = FALSE) +
  geom_text(aes(label = Count), vjust = -0.5, size = 4, fontface = "bold") +
  labs(title = "Missing Value Analysis by Column",
       x = "", y = "Number of Columns") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```
As seen in the bar graph, from a total of **763 features, 242 features** have zero missing value entries. There are **496** features having a low number of missing values ranging from 0 to 10% and **25 features** containing a medium number of missing values ranging from 10 to 50%. The good thing is none of the features has greater than **50% missing values**. We will be handling the missing values using median imputation.

## Data Preprocessing
At this level, we shall be undertaking the process of cleaning our dataset. This is because raw data cannot be used to input into our machine learning algorithms. These are the three main processes that we undertake; these include the elimination of features whose variance is zero, the filling up of the missing data using median imputation, and the elimination of highly correlated features.

```{r preprocessing, message=FALSE, warning=FALSE}
# Create binary target variable
processed_data <- raw_data %>%
  mutate(
    target = factor(ifelse(loss > 0, 1, 0)),
    loss = loss / 100
  )

cat("Class distribution:\n")
print(table(processed_data$target))

# Remove near zero variance features
feature_cols <- setdiff(names(processed_data), c("id", "loss", "target"))
nzv_indices <- nearZeroVar(processed_data[, feature_cols])

if (length(nzv_indices) > 0) {
  processed_data <- processed_data[, !names(processed_data) %in% feature_cols[nzv_indices]]
  cat(sprintf("Removed %d near-zero variance features\n", length(nzv_indices)))
}

# Median imputation and correlation removal
feature_cols2 <- setdiff(names(processed_data), c("id", "loss", "target"))
preproc <- preProcess(processed_data[, feature_cols2],
                      method = c("medianImpute", "corr"))
processed_clean <- predict(preproc, processed_data)
processed_clean$target <- processed_data$target

cat(sprintf("Features remaining after preprocessing: %d\n",
            ncol(processed_clean) - 3))
```
The feature space was compressed from **763 dimensions down to 246**. There were **24 features with near-zero** variance that did not contribute to the predictive power of the model, which were therefore dropped. The missing data was dealt with using median imputation, and highly correlated variables were dropped to **avoid multicollinearity issues**.

## Feature Selection — LASSO
We take help from LASSO regression in this step to narrow down the top features from the large pool of 246 features. In fact, LASSO shrinks the coefficients of less important features to zero which results in their elimination. As a result, we get a smaller and more targeted set of features that are primarily linked to loan default prediction.

```{r lasso, message=FALSE, warning=FALSE}
feature_cols3 <- setdiff(names(processed_clean), c("id", "loss", "target"))
x_cls <- data.matrix(processed_clean[, feature_cols3])
y_cls <- as.numeric(as.character(processed_clean$target))

set.seed(123)
lasso_cls <- cv.glmnet(x_cls, y_cls,
                        alpha = 1,
                        family = "binomial",
                        nfolds = 10,
                        type.measure = "auc")

plot(lasso_cls, main = "LASSO Cross Validation — Feature Selection")

coef_cls <- coef(lasso_cls, s = "lambda.min")
lasso_features <- data.frame(
  name = coef_cls@Dimnames[[1]][coef_cls@i + 1],
  coefficient = abs(coef_cls@x)
) %>%
  filter(name != "(Intercept)") %>%
  arrange(desc(coefficient)) %>%
  pull(name) %>%
  as.vector()

cat(sprintf("LASSO selected %d features\n", length(lasso_features)))

lasso_data <- processed_clean %>%
  select(all_of(c(lasso_features, "target")))
```
 
A **LASSO cross-validation** method is carried out using 10-folds to identify the top features to predict loan default. As depicted in the graph below, the AUC score rises with an increasing number of features until it reaches a stable point at around **0.70**. This implies that additional features do not help after a particular threshold. LASSO identified **180** features out of 246 features through **coefficient shrinkage**.

## Model 1 — Random Forest Classifier

At this stage, we develop a Random Forest classifier to predict if there will be any default on the loan. A Random Forest algorithm creates many decision trees and then merges their predictions. The selection of Random Forest is because of its ability to work on imbalanced datasets, identify non-linear interactions, and provide the importance of features. Due to the imbalance in our dataset where defaults are 9.22%, class weight is used.

```{r random-forest, message=FALSE, warning=FALSE}
set.seed(123)

# Train/Validation split
rf_index <- createDataPartition(lasso_data$target, p = 0.80, list = FALSE)
rf_train <- lasso_data[rf_index, ]
rf_validate <- lasso_data[-rf_index, ]

# Train Random Forest
rf_model <- randomForest(
  target ~ .,
  data = rf_train,
  ntree = 300,
  mtry = floor(sqrt(ncol(rf_train) - 1)),
  importance = TRUE,
  classwt = c("0" = 1, "1" = 5)
)

print(rf_model)
```
The random forest has been created with the use of **300 decision trees**, where each node has tested 13 variables. The out-of-bag **error rate is 9.26%**, which seems relatively low; however, it does not truly reflect the performance of the model due to the unbalanced classes. It is observed from the confusion matrix that **58,063** observations have been correctly classified as non-defaulter cases, and out of **5,904 defaulter cases, 10 have been correctly identified**. The model has performed slightly better than before by giving class weight of **5 to the defaulter cases**.

### ROC Curve

ROC curve helps measure the performance of our model at different thresholds. AUC helps measure the ability of our model to differentiate between default and non-default loans. If our AUC value equals 0.5, then we have no model performance since that value corresponds to random guessing, while an AUC equaling 1.0 indicates perfect prediction by the model.

```{r roc-curve, message=FALSE, warning=FALSE}
# Get predicted probabilities
rf_probs <- predict(rf_model, newdata = rf_validate, type = "prob")

# ROC curve
roc_obj <- roc(as.numeric(as.character(rf_validate$target)),
               rf_probs[, "1"], quiet = TRUE)

# Plot
plot(roc_obj,
     main = paste0("ROC Curve — Random Forest (AUC = ", round(auc(roc_obj), 4), ")"),
     col = "#5C9BE0", lwd = 2)
abline(a = 0, b = 1, lty = 2, col = "grey60")
legend("bottomright",
       legend = c(paste0("AUC = ", round(auc(roc_obj), 4)), "Random Baseline"),
       col = c("#5C9BE0", "grey60"),
       lty = c(1, 2), lwd = c(2, 1), bty = "n")

cat(sprintf("AUC Score: %.4f\n", auc(roc_obj)))
```
According to the ROC curve analysis, the Random Forest classification algorithm has managed to reach an **AUC value of 0.6212**. Therefore, we can conclude that the model has some capacity to discriminate between defaulting and non-defaulting loans, although there is still some room for improvement considering that the value is greater than 0.5.

## Model 2 — XGBoost Regression

Here, we will create an XGBoost model to predict the loss amount of loans expected to be defaulters. XGBoost stands for eXtreme Gradient Boosting and is a machine learning model based on gradient boosting which involves building trees sequentially such that each tree corrects the error made by its predecessor. The XGBoost model will only be trained on defaulted loans since we only have to predict the loss amount in case of defaults. XGBoost models are preferred because they can capture non-linearity, are not sensitive to outliers and perform well on financial datasets.

```{r xgboost, message=FALSE, warning=FALSE}
# Filter only defaulted loans for regression
regression_train <- raw_data %>%
  filter(loss > 0) %>%
  mutate(loss = loss / 100)

# Preprocessing
feature_cols_reg <- setdiff(names(regression_train), c("id", "loss"))
nzv_reg <- nearZeroVar(regression_train[, feature_cols_reg])
if (length(nzv_reg) > 0) {
  regression_train <- regression_train[, !names(regression_train) %in% feature_cols_reg[nzv_reg]]
}

feature_cols_reg2 <- setdiff(names(regression_train), c("id", "loss"))
reg_prep <- preProcess(regression_train[, feature_cols_reg2],
                       method = c("medianImpute", "corr"))
regression_clean <- predict(reg_prep, regression_train)
regression_clean$loss <- regression_train$loss

# LASSO feature selection
feature_cols_reg3 <- setdiff(names(regression_clean), c("id", "loss"))
x_reg <- as.matrix(regression_clean[, feature_cols_reg3])
y_reg <- regression_clean$loss

set.seed(123)
lasso_reg <- cv.glmnet(x_reg, y_reg, alpha = 1, family = "gaussian",
                        nfolds = 10, type.measure = "mse")

coef_reg <- coef(lasso_reg, s = "lambda.min")
lasso_feat_reg <- data.frame(
  name = coef_reg@Dimnames[[1]][coef_reg@i + 1],
  coefficient = abs(coef_reg@x)
) %>%
  filter(name != "(Intercept)") %>%
  arrange(desc(coefficient)) %>%
  pull(name) %>%
  as.vector()

# Train/Validation split
reg_data <- regression_clean %>% select(all_of(c(lasso_feat_reg, "loss")))
set.seed(123)
reg_index <- createDataPartition(reg_data$loss, p = 0.80, list = FALSE)
reg_train <- reg_data[reg_index, ]
reg_val <- reg_data[-reg_index, ]

x_train <- as.matrix(reg_train[, lasso_feat_reg])
y_train <- reg_train$loss
x_val <- as.matrix(reg_val[, lasso_feat_reg])
y_val <- reg_val$loss

dtrain <- xgb.DMatrix(data = x_train, label = y_train)
dval <- xgb.DMatrix(data = x_val, label = y_val)

set.seed(123)
xgb_model <- xgb.train(
  params = list(
    objective = "reg:squarederror",
    eval_metric = "mae",
    learning_rate = 0.03,
    max_depth = 4,
    subsample = 0.8,
    colsample_bytree = 0.8
  ),
  data = dtrain,
  nrounds = 500,
  watchlist = list(val = dval),
  early_stopping_rounds = 30,
  verbose = 0
)

# Plot 1 — Training Progress
eval_log <- as.data.frame(attr(xgb_model, "evaluation_log"))
ggplot(eval_log, aes(x = iter, y = val_mae)) +
  geom_line(color = "#5C9BE0", linewidth = 1) +
  labs(title = "XGBoost — Training Progress",
       x = "Round", y = "Validation MAE") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

# Plot 2 — Predicted vs Actual
xgb_preds <- predict(xgb_model, dval)
xgb_preds_rescaled <- pmax(xgb_preds * 100, 0)
y_val_rescaled <- y_val * 100

plot_df <- data.frame(actual = y_val_rescaled, predicted = xgb_preds_rescaled)

ggplot(plot_df, aes(x = actual, y = predicted)) +
  geom_point(alpha = 0.3, color = "#5C9BE0", size = 1.2) +
  geom_abline(slope = 1, intercept = 0, color = "#E05C5C", linetype = "dashed", linewidth = 1) +
  labs(title = "XGBoost — Predicted vs Actual Loss",
       x = "Actual Loss", y = "Predicted Loss") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

# Plot 3 — Feature Importance
xgb_importance <- xgb.importance(model = xgb_model)
xgb.plot.importance(xgb_importance[1:min(20, nrow(xgb_importance)), ],
                    main = "XGBoost — Top 20 Feature Importances")
```
### Plot 1 – Training Progress:
The plot of training progress indicates an initially quick drop in validation MAE that gradually stabilizes as the algorithm converges. With a reduced **learning rate of 0.03**, the model takes a little longer to learn but with higher precision when compared to the previous parameters. Early stopping made sure there was no overfitting through optimal rounds.

### Plot 2 – Predicted vs Actual:
From the comparison plot, the accuracy of the model is improved from the previous model. There are more clustered data points close to the **red dashed line**. This model is however challenged with **high-loss figures above 50**. These are rare occurrences in the training dataset.

### Plot 3 – Feature Importance:

This plot is consistent with the previous plot on feature importance where **f766** continues being the most dominant feature while **f67 and f229** come second and third respectively. However, the spread of importance score is wider in this new model.

## Model Evaluation

This section will discuss the performance of both models. The metrics for the random forest classifier include accuracy, precision, recall, f1 score, and AUC while those of the xgboost regressor include MAE and RMSE.

```{r evaluation, message=FALSE, warning=FALSE}
# ── Random Forest Evaluation ──────────────────────────────────────
rf_probs <- predict(rf_model, newdata = rf_validate, type = "prob")
rf_pred_adjusted <- factor(ifelse(rf_probs[, "1"] >= 0.068, "1", "0"), 
                           levels = c("0", "1"))

cm <- confusionMatrix(rf_pred_adjusted, rf_validate$target, positive = "1")

accuracy  <- cm$overall["Accuracy"]
precision <- cm$byClass["Pos Pred Value"]
recall    <- cm$byClass["Sensitivity"]
f1        <- 2 * (precision * recall) / (precision + recall)
roc_obj   <- roc(as.numeric(as.character(rf_validate$target)),
                 rf_probs[, "1"], quiet = TRUE)
auc_score <- auc(roc_obj)

cat("═══════════════════════════════════════\n")
cat(" RANDOM FOREST — CLASSIFICATION METRICS\n")
cat("═══════════════════════════════════════\n")
cat(sprintf(" Accuracy  : %.4f\n", accuracy))
cat(sprintf(" Precision : %.4f\n", precision))
cat(sprintf(" Recall    : %.4f\n", recall))
cat(sprintf(" F1 Score  : %.4f\n", f1))
cat(sprintf(" AUC       : %.4f\n", auc_score))
cat("═══════════════════════════════════════\n")

# ── XGBoost Evaluation ────────────────────────────────────────────
xgb_preds <- predict(xgb_model, dval)
xgb_preds_rescaled <- pmax(xgb_preds * 100, 0)
y_val_rescaled <- y_val * 100

mae  <- mean(abs(xgb_preds_rescaled - y_val_rescaled))
rmse <- sqrt(mean((xgb_preds_rescaled - y_val_rescaled)^2))

cat("\n═══════════════════════════════════════\n")
cat(" XGBOOST — REGRESSION METRICS\n")
cat("═══════════════════════════════════════\n")
cat(sprintf(" MAE  : %.4f\n", mae))
cat(sprintf(" RMSE : %.4f\n", rmse))
cat("═══════════════════════════════════════\n")

# Plot 1 — Confusion Matrix Heatmap
cm_table <- as.data.frame(cm$table)
ggplot(cm_table, aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(color = "white", linewidth = 1) +
  geom_text(aes(label = Freq), size = 7, fontface = "bold", color = "white") +
  scale_fill_gradient(low = "#5C9BE0", high = "#E05C5C") +
  labs(title = "Random Forest — Confusion Matrix (Threshold = 0.068)",
       x = "Actual", y = "Predicted") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        legend.position = "none")

# Plot 2 — Residual Plot
plot_df <- data.frame(actual = y_val_rescaled, predicted = xgb_preds_rescaled)
plot_df$residual <- plot_df$predicted - plot_df$actual

ggplot(plot_df, aes(x = actual, y = residual)) +
  geom_point(alpha = 0.2, color = "#5C9BE0", size = 1) +
  geom_hline(yintercept = 0, color = "#E05C5C", linetype = "dashed", linewidth = 1) +
  labs(title = "XGBoost — Residual Plot",
       subtitle = "Points above zero = overprediction | Points below zero = underprediction",
       x = "Actual Loss", y = "Residual (Predicted - Actual)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, color = "grey50"))
```

### Metrics:
Random forest classifier gave us the **accuracy of 56.65%** with **recall equal to 60.95%**, and we have increased the capability of our model to detect defaults by about 61%. The reason for this significant improvement was adjusting classification threshold to **0.068** by Youden's method and then increasing the class weight up to **5**. **F1 score equal to 0.2059 and AUC score of 0.6212** show average results on the whole. As for the XGBoost classifier, the improvement came due to adjusting such hyperparameters as learning rate to 0.03 and max_depth to 4 with the result of **MAE = 5.2261 and RMSE = 10.4407**.

### Confusion Matrix:
The confusion matrix gives us information that out of **1475 actual defaults**, **899** were correctly detected by our algorithm, but **576 went undetected**. In terms of non-defaults, our algorithm classified 8165 samples correctly as opposed to 6359 that were falsely detected as defaults.

### Residual Plot:
The residual plot shows that for **small loss values** the residuals are mostly close to zero indicating good predictions. For **larger actual loss** values the residuals become increasingly negative meaning the model underpredicts extreme losses. This pattern is consistent with the **right skewed nature of our loss distribution**.

## Predictions on Test Data

In this stage, we use the trained models on the testing data to produce our loss predictions. Our pipeline remains identical to the previous stage, consisting of the first step, where Random Forest classifies default vs. non-default of loans, and the second step, where XGBoost predicts the amount of loss given that the loan defaults.

```{r predictions, message=FALSE, warning=FALSE}
# Load test data
test_data <- read_csv(TEST_PATH, show_col_types = FALSE)

# ── Step 1: Preprocess test data same as training ─────────────────
test_processed <- predict(preproc, test_data)

# Align features with LASSO selected classification features
test_cls_feats <- intersect(lasso_features, names(test_processed))
test_cls_data <- test_processed[, test_cls_feats, drop = FALSE]

# ── Step 2: Classify defaults using Random Forest ─────────────────
test_probs <- predict(rf_model, newdata = test_cls_data, type = "prob")
test_default_flag <- ifelse(test_probs[, "1"] >= 0.068, 1, 0)

cat(sprintf("Predicted defaults     : %d (%.1f%%)\n",
            sum(test_default_flag),
            mean(test_default_flag) * 100))
cat(sprintf("Predicted non-defaults : %d (%.1f%%)\n",
            sum(test_default_flag == 0),
            mean(test_default_flag == 0) * 100))

# ── Step 3: Predict loss using XGBoost ───────────────────────────
test_reg_processed <- predict(reg_prep, test_data)
test_reg_feats <- intersect(lasso_feat_reg, names(test_reg_processed))
test_reg_data <- as.matrix(test_reg_processed[, test_reg_feats, drop = FALSE])

dtest <- xgb.DMatrix(data = test_reg_data)
test_loss_pred <- pmax(predict(xgb_model, dtest) * 100, 0)

# ── Step 4: Combine predictions ───────────────────────────────────
final_loss <- ifelse(test_default_flag == 1, test_loss_pred, 0)

cat(sprintf("\nFinal predictions summary:\n"))
cat(sprintf("Rows with loss = 0 : %d\n", sum(final_loss == 0)))
cat(sprintf("Rows with loss > 0 : %d\n", sum(final_loss > 0)))
cat(sprintf("Min loss : %.4f\n", min(final_loss)))
cat(sprintf("Max loss : %.4f\n", max(final_loss)))
cat(sprintf("Mean loss: %.4f\n", mean(final_loss)))
```
This **two-pipeline model** was used on a test data sample consisting of **25,471 loan applications**. The classification using the **Random Forest** algorithm predicted **11,424 loan defaults**, which constitutes **44.9%** of the test sample. The loss values for the predicted loan defaults were calculated by the **XGBoost** model, varying between **0 and 54.19** with a mean loss of **3.69**. Loan applications not falling in the ‘default’ category were tagged a value of **zero loss**.

## Saving the Predictions File

In this final step we save our predictions to a CSV file in the required format with two columns — id and loss.
```{r save-predictions, message=FALSE, warning=FALSE}
# Build submission file
submission <- data.frame(
  id   = test_data$id,
  loss = round(final_loss, 4)
)

# Validation checks
stopifnot("Wrong row count" = nrow(submission) == nrow(test_data))
stopifnot("Negative losses" = all(submission$loss >= 0))

# Preview
cat("Submission file preview:\n")
print(head(submission, 10))

cat(sprintf("\nTotal rows : %d\n", nrow(submission)))
cat(sprintf("Loss > 0   : %d\n", sum(submission$loss > 0)))
cat(sprintf("Loss = 0   : %d\n", sum(submission$loss == 0)))

# Save
write.csv(submission, "final_predictions.csv", row.names = FALSE)
cat("\nfinal_predictions.csv saved successfully!\n")
```
The file to be submitted has been created successfully containing 25,471 rows with the id and loss fields as specified. Of the total of 25,471 loans, **11,421** had a positive loss amount while the rest of **14,050** were assigned a zero loss value meaning that there was no default on the loan.

## Conclusion

In this project, we sought to use machine learning algorithms to predict the probability of loan defaulting and the estimated financial loss in case of defaulting. We employed a two-step approach where in step one, we applied a Random Forest classifier to determine whether a loan would default or not. In the second step, we used an XGBoost regression algorithm to predict the loss amount in case of loan default.

Major findings from this project:

- The dataset had an imbalance ratio of 9.22%, making it difficult for the classifier to perform optimally
- Despite tuning the class weights and classification threshold, the Random Forest classifier obtained a recall score of 60.95% which indicated that the classifier managed to detect about 61% of loan defaults
- The optimized XGBoost model attained a mean absolute error (MAE) of 5.2261 against the initial MAE of 5.4467, which was an improvement
- f766 is the most significant predictor of loan loss amount, followed by f67 and f229
- There is a challenge of predicting losses beyond 50 because of the scarcity of such losses in the training dataset

Future research can focus on trying more sophisticated ensembling techniques, gathering more rare loss cases data, and implementing deep learning techniques.

